---
title: "Ch 3"
author: "Britt"
date: '2022-08-30'
output: html_document
---
run lobstr

```{r}
library(lobstr)
knitr::opts_chunk$set(echo = TRUE)
```

Ch. 3: Vectors

3.1 Introduction

atomic vectors: all elements must have the same type
lists: elements can have different types
NULL: not a vector, general "zero length" vector

attribute: named list of arbitrary metadata
-every vector can have attributes

two important attributes are dimension and class
dimension: turns vectors into matrices and arrays
class: powers S3 object system
- important S3 vectors: factors, date and times, data frames, and tibbles

3.2 Atomic Vectors

four primary types of atomic vectors: logical, integer, double, and character (which contains strings)
-integer and double vectors are both numeric vectors
-complex and raw vectors are rare and needed only in certain cases

3.2.1 Scalars

scalar: special syntax to create an individual value

Logicals: written in full (TRUE or FALSE), or abbreviated (T or F).
Doubles: specified in decimal (0.1234), scientific (1.23e4), or hexadecimal (0xcafe) form. There are three special values unique to doubles: Inf, -Inf, and NaN (not a number). These are special values defined by the floating point standard.
Integers: written similarly to doubles but must be followed by L23 (1234L, 1e4L, or 0xcafeL), and can not contain fractional values.
Strings: surrounded by " ("hi") or ' ('bye'). Special characters are escaped with \; see ?Quotes for full details.

3.2.2 Making Longer Vectors with c()

c(): short for combine, makes longer vectors from shorter ones

When the inputs are atomic vectors, c() always creates another atomic vector; i.e. it flattens:

```{r}
lgl_var <- c(TRUE, FALSE)
int_var <- c(1L, 6L, 10L)
dbl_var <- c(1, 2.5, 4.5)
chr_var <- c("these are", "some strings")
```
Kyle's example of c() flattening:
```{r}
lgl_var <- c(TRUE, FALSE)
int_nested_var <- c(c(1L,2L), c(3L,4L))
obj_addr(c(1L, 2L))
obj_addr(c(3L, 4L))
obj_addr(c(c(1L, 2L), c(3L, 4L)))
typeof(lgl_var)
typeof(int_nested_var)
```
c() puts the vectors in a new location where they are both located, rather than two separate locations

use tpeof() to determine the type of vector and length() to determine its length:
```{r}
typeof(lgl_var)
#> [1] "logical"
typeof(int_var)
#> [1] "integer"
typeof(dbl_var)
#> [1] "double"
typeof(chr_var)
#> [1] "character"
```
3.2.3 Missing Values

NA: missing or unknown values
-most computations involving a missing value will return another missing value (be careful)

```{r}
NA > 5
#> [1] NA
10 * NA
#> [1] NA
!NA
#> [1] NA
```
only a few exceptions to this rule, which occur when some identity holds for all possible inputs:

```{r}
NA ^ 0
#> [1] 1
NA | TRUE
#> [1] TRUE
NA & FALSE
#> [1] FALSE
```
Propagation of missingness leads to a common mistake when determining which values in a vector are missing:

```{r}
x <- c(NA, 5, NA, 10)
x == NA
#> [1] NA NA NA NA
```
This result is correct (if a little surprising) because there’s no reason to believe that one missing value has the same value as another. Instead, use is.na() to test for the presence of missingness:

```{r}
is.na(x)
#> [1]  TRUE FALSE  TRUE FALSE
```
3.2.4 Testing and Coercion

You can test if a vector is a given type with an is.*() function. 
ex: is.logical(), is.integer(), is.double(), is.character(), but not is.vector(), is.atomic(), and is.numeric().

As a reminder, atomic vectors must have elements all of the same type. If they are not the same type, they will be coerced in a fixed order: character --> double --> integer --> logical. For example, combining a character and an integer yields a character:

```{r}
str(c("a", 1))
#>  chr [1:2] "a" "1"
```
Character is first in the order of vectors that will be coerced when mixed elements are used in an atomic vector. This is why the above results in a character even though an integer was also combined with it. 

Coercion happens automatically.

Most math functions (ex: +, log, abs, etc) will coerce to numeric. This is useful for logical vectors because true becomes 1 and false becomes 0.

```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)
#> [1] 0 0 1

# Total number of TRUEs
sum(x)
#> [1] 1

# Proportion that are TRUE
mean(x)
#> [1] 0.333
```
You can deliberately coerce by using an as.*() function. Failed coercion of strings creates a warning and missing value:

```{r}
as.integer(c("1", "1.5", "a"))
#> Warning: NAs introduced by coercion
#> [1]  1  1 NA
```
3.2.5 Exercises

1. How do you create raw and complex scalars?

```{r}
?raw
?complex
```

2. Test your knowledge of the vector coercion rules by predicting the output of the following uses of c():

c(1, FALSE) --> 1, 0 (logical has less priority than double, so coercion makes them both #s)
c("a", 1) --> "a", "1" (character has top priority for coercion)
c(TRUE, 1L) --> 1, 1 (logical has less priority than integer)

```{r}
c(1, FALSE)
c("a", 1)
c(TRUE, 1L)
```
3. Why is 1 == "1" true? Why is -1 < FALSE true? Why is "one" < 2 false?
1 will be coerced into "1" since character takes priority. "1" = "1"
FALSE will be coerced into 0 since math vectors change logical to integer. -1 < 0
2 will be coerced to "2" since character takes priority, but "one" /</ "2" (these are characters not numbers!)

```{r}
1 == "1"
-1 < FALSE
"one" < 2
```
4. Why is the default missing value, NA, a logical vector? What's special about logical vectors? (Hint: think about c(FALSE, NA_character_))

Logical vectors have the least priority in coercion. NULL is also essentially 0, or a vector with no length, and in this sense, is most similar to FALSE/0, as in a logical vector. I'm not sure what is special about logical vectors other than the fact that they only have 1/0 and TRUE/FALSE options.

```{r}
c(FALSE, NA_character_) 
```
5. Precisely what do is.atomic(), is.numeric(), and is.vector() test for?
is.atomic, is.numeric, and is.vector test to see if the given vector is of the specified mode with only names. TRUE = this is true, FALSE = anything other than the correct mode with only names.

3.3 Attributes

-data structures like matrices, arrays, factors, date-times, etc are built on top of atomic vectors by adding attributes

3.3.1 Getting and Setting

attributes: name-value pairs that attach metadata to an object
attr(): retrieves and modifies individual attributes
attributes(): retrieves attribues en masse
structure(): sets attributes en masse

```{r}
a <- 1:3
attr(a, "x") <- "abcdef"
attr(a, "x")
#> [1] "abcdef"

attr(a, "y") <- 4:6
str(attributes(a))
#> List of 2
#>  $ x: chr "abcdef"
#>  $ y: int [1:3] 4 5 6

# Or equivalently
a <- structure(
  1:3, 
  x = "abcdef",
  y = 4:6
)
str(attributes(a))
#> List of 2
#>  $ x: chr "abcdef"
#>  $ y: int [1:3] 4 5 6
```
Most attributes are lost by most operations:
```{r}
attributes(a[1])
#> NULL
attributes(sum(a))
#> NULL
```
names: a character vector giving each element a name, an attribute that is routinely preserved
dim: (short for dimensions) an integer vector that is used to turn vectors into matrices or arrays, an attribute that is routinely preserved
-other attributes must be preserved by creating your own S3 class (ch. 13)

3.3.2 Names

There are three ways to name a vector.

```{r}
# When creating it: 
x <- c(a = 1, b = 2, c = 3)

# By assigning a character vector to names()
x <- 1:3
names(x) <- c("a", "b", "c")

# Inline, with setNames():
x <- setNames(1:3, c("a", "b", "c"))
```

Avoid using attr(x, "names"). It requires more typing and is less readable than names(x).
You can remove names from a vector by using x <- unname(x) or names(x) <- NULL.
Names should be unique and non-missing.

3.3.3 Dimensions

-Adding a dim attribute to a vector allows it to behave like a 2-dimensional matrix or multi-dimensional array. Matrices and arrays can be created with matrix() and array() by using the assignment form of dim().

```{r}
# Two scalar arguments specify row and column sizes
x <- matrix(1:6, nrow = 2, ncol = 3)
x
#>      [,1] [,2] [,3]
#> [1,]    1    3    5
#> [2,]    2    4    6

# One vector argument to describe all dimensions
y <- array(1:12, c(2, 3, 2))
y
#> , , 1
#> 
#>      [,1] [,2] [,3]
#> [1,]    1    3    5
#> [2,]    2    4    6
#> 
#> , , 2
#> 
#>      [,1] [,2] [,3]
#> [1,]    7    9   11
#> [2,]    8   10   12

# You can also modify an object in place by setting dim()
z <- 1:6
dim(z) <- c(3, 2)
z
#>      [,1] [,2]
#> [1,]    1    4
#> [2,]    2    5
#> [3,]    3    6
```
A vector without a dim attribute is often thought of as 1-dimensional, but actually has NULL dimensions. You also can have matrices with a single row or single column, or arrays with a single dimension. They may print similarly, but will behave differently. The differences aren’t too important, but it’s useful to know they exist in case you get strange output from a function (tapply() is a frequent offender). As always, use str() to reveal the differences.

```{r}
str(1:3)                   # 1d vector
#>  int [1:3] 1 2 3
str(matrix(1:3, ncol = 1)) # column vector
#>  int [1:3, 1] 1 2 3
str(matrix(1:3, nrow = 1)) # row vector
#>  int [1, 1:3] 1 2 3
str(array(1:3, 3))         # "array" vector
#>  int [1:3(1d)] 1 2 3
```
3.3.4 Exercises

1. How is setNames() implemented? How is unname() implemented? Read the source code. 

setNames() helps set names after entering data for convenience. Unname() can unname objects in case of mistake or otherwise.

2. What does dim() return when applied to a 1-dimensional vector? When might you use NROW() or NCOL()?

```{r}
dim(c(1,2,3))
```
The output is NULL (dimensions are only valid at the 2D level essentially).
NROW and NCOL would be used at 2D or higher.

3. How would you describe the following three objects? What makes them different from 1:5?

```{r}
x1 <- array(1:5, c(1, 1, 5))
x2 <- array(1:5, c(1, 5, 1))
x3 <- array(1:5, c(5, 1, 1))
```

They differ by which row has the 5 digit space in it.

4. An early draft used this code to illustrate structure():
```{r}
structure(1:5, comment = "my attribute")
#> [1] 1 2 3 4 5

?comment
```
But when you print that object you don't see the comment attribute. Why? Is the attribute missing, or is there something else special about it? (Hint: try using help)

Comment is used for a human comment (a note to read later), but not something that can be used in the code.

3.4 S3 Atomic Vectors

-one of the most important vector attributes is class
-having a class attribute turns an oject into an S3 object

Four important S3 vectors used in base R:
categorical data, where values come from a fixed set of levels recorded in factor vectors.
Dates (with day resolution), which are recorded in Date vectors.
Date-times (with second or sub-second resolution), which are stored in POSIXct vectors.
Durations, which are stored in difftime vectors.

3.4.1 Factors

factor: a vector that can only contain predefined values, used to store categorical data. They are built on top of an integer vector with two attributes: class and levels.

```{r}
x <- factor(c("a", "b", "b", "a"))
x
#> [1] a b b a
#> Levels: a b

typeof(x)
#> [1] "integer"
attributes(x)
#> $levels
#> [1] "a" "b"
#> 
#> $class
#> [1] "factor"
```
Factors are useful when you know the set of possible values but they're not all present in a given dataset. In contrast to a character vector, when you tabulate a factor you'll get counts of all categories, even unobserved ones.

```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
#> sex_char
#> m 
#> 3
table(sex_factor)
#> sex_factor
#> m f 
#> 3 0
```
Ordered factors: behave like regular factors, but the order of the levels is meaningful (low, med, high)

```{r}
grade <- ordered(c("b", "b", "a", "c"), levels = c("c", "b", "a"))
grade
#> [1] b b a c
#> Levels: c < b < a
```
-in base R, you tend to encounter factors frequently because many functions (like read.csv() and data.frame()) automatically convert character vectors to factors. Use the argument stringsAsFactors = FALSE to suppress this behaviour, and then manually convert character vectors to factors using your knowledge of the “theoretical” data. 
-factors look and often behave like character vectors, but are built on top of integers.

3.4.2 Dates

Date vectors are built on top of double vectors. They have class "date" and no other attributes.

```{r}
today <- Sys.Date()

typeof(today)
#> [1] "double"
attributes(today)
#> $class
#> [1] "Date"
```
The value of the double (which can be seen by stripping the class), represents the number of days since 1970-01-01.

```{r}
date <- as.Date("1970-02-01")
unclass(date)
#> [1] 31
```
3.4.3 Date-Times

Two ways of storing date-time information: POSIXct and POSIXIt. 

```{r}
now_ct <- as.POSIXct("2018-08-01 22:00", tz = "UTC")
now_ct
#> [1] "2018-08-01 22:00:00 UTC"

typeof(now_ct)
#> [1] "double"
attributes(now_ct)
#> $class
#> [1] "POSIXct" "POSIXt" 
#> 
#> $tzone
#> [1] "UTC"
```
The tzone attribue controls only how the date-time is formatted. It does not control the instant of time represented by the vector. Note that the time is not printed if it is midnight.

```{r}
structure(now_ct, tzone = "Asia/Tokyo")
#> [1] "2018-08-02 07:00:00 JST"
structure(now_ct, tzone = "America/New_York")
#> [1] "2018-08-01 18:00:00 EDT"
structure(now_ct, tzone = "Australia/Lord_Howe")
#> [1] "2018-08-02 08:30:00 +1030"
structure(now_ct, tzone = "Europe/Paris")
#> [1] "2018-08-02 CEST"
```
3.4.4 Durations

duration: amount of time between pairs of dates or date-times, stored in difftimes
difftimes: built on top of doubles and have a units attribute

```{r}
one_week_1 <- as.difftime(1, units = "weeks")
one_week_1
#> Time difference of 1 weeks

typeof(one_week_1)
#> [1] "double"
attributes(one_week_1)
#> $class
#> [1] "difftime"
#> 
#> $units
#> [1] "weeks"

one_week_2 <- as.difftime(7, units = "days")
one_week_2
#> Time difference of 7 days

typeof(one_week_2)
#> [1] "double"
attributes(one_week_2)
#> $class
#> [1] "difftime"
#> 
#> $units
#> [1] "days"
```
3.4.5 Exercises

1. What sort of object does table() return? What is its type? What attributes does it have? How does the dimensionality change as you tabulate more variables?

```{r}
typeof(table("a", "b"))
attributes(table("a","b", "c"))

attributes(table(c(1,2), c(1,2), c(1,2)))
```
Table() returns an integer. Attributes are dimensions and class. As you tabulate more objects, it increases in rows. 1, then 1 1, then 1 1 1, etc.

2. What happens to a factor when you modify its levels? ?
```{r}
f1 <- factor(letters)
levels(f1) <- rev(levels(f1))
```

3. What does this code do? How do f2 and f3 differ from f1? ?
```{r}
f2 <- rev(factor(letters))

f3 <- factor(letters, levels = rev(letters))

f1
f2
f3
```

3.5 Lists

lists: a step up in complexity from atomic vectors: each element can be any type, not just vectors. Technically speaking, each element of a list is actually the same type because, as you saw in Section 2.3.3, each element is really a reference to another object, which can be any type.

3.5.1 Creating

list(): construct a list

```{r}
l1 <- list(
  1:3, 
  "a", 
  c(TRUE, FALSE, TRUE), 
  c(2.3, 5.9)
)

typeof(l1)
#> [1] "list"

str(l1)
#> List of 4
#>  $ : int [1:3] 1 2 3
#>  $ : chr "a"
#>  $ : logi [1:3] TRUE FALSE TRUE
#>  $ : num [1:2] 2.3 5.9
```
Because the elements of a list are references, creating a list does not involve copying the components into the list. For this reason, the total size of a list might be smaller than you might expect.

```{r}
lobstr::obj_size(mtcars)
#> 7,208 B

l2 <- list(mtcars, mtcars, mtcars, mtcars)
lobstr::obj_size(l2)
#> 7,288 B
```
Lists are sometimes called recursive vectors because a list can contain other lists. This makes them fundamentally different from atomic vectors.

```{r}
l3 <- list(list(list(1)))
str(l3)
#> List of 1
#>  $ :List of 1
#>   ..$ :List of 1
#>   .. ..$ : num 1
```
c() will combine several lists into one. If given a combination of atomic vectors and lists, c() will coerce the vectors to lists before combining them. Compare the results of list() and c():

```{r}
l4 <- list(list(1, 2), c(3, 4))
l5 <- c(list(1, 2), c(3, 4))
str(l4)
#> List of 2
#>  $ :List of 2
#>   ..$ : num 1
#>   ..$ : num 2
#>  $ : num [1:2] 3 4
str(l5)
#> List of 4
#>  $ : num 1
#>  $ : num 2
#>  $ : num 3
#>  $ : num 4
```
3.5.2 Testing and Coercion

is.list(): test for a list
as.list(): coerce to a list

```{r}
list(1:3)
#> [[1]]
#> [1] 1 2 3
as.list(1:3)
#> [[1]]
#> [1] 1
#> 
#> [[2]]
#> [1] 2
#> 
#> [[3]]
#> [1] 3
```
unlist(): turn a list into an atomic vector

3.5.3 Matrices and Arrays

With atomic vectors, the dimension attribute is commonly used to create matrices. With lists, the dimension attribute can be used to create list-matrices or list-arrays:

```{r}
l <- list(1:3, "a", TRUE, 1.0)
dim(l) <- c(2, 2)
l
#>      [,1]      [,2]
#> [1,] Integer,3 TRUE
#> [2,] "a"       1

l[[1, 1]]
#> [1] 1 2 3
```
These data structures are relatively esoteric but they can be useful if you want to arrange objects in a grid-like structure. For example, if you’re running models on a spatio-temporal grid, it might be more intuitive to store the models in a 3D array that matches the grid structure.

3.5.4 Exercises

1. List all the ways that a list differs from an atomic vector. A list can have different types, but an atomic vector must have all the same types. A list is references to objects, but an atomic vector is the actual objects.

2. Why do you need to use unlist() to convert a list to an atomic vector? Why doesn't as.vector() work? ?

```{r}
l1 <- list(
    1:3,
    "a",
    c(T, F, T),
    c(2.3, 5.9)
)

unlist(l1)
as.vector(l1)
```
3. Compare and contrast c() and unlist() when combining a dat and date-time into a single vector. ?

3.6 Data Frames and Tibbles

Two most important S3 vectors built on top of lists: data frames and tibbles.

If you do data analysis in R, you’re going to be using data frames. A data frame is a named list of vectors with attributes for (column) names, row.names29, and its class, “data.frame”:

```{r}
df1 <- data.frame(x = 1:3, y = letters[1:3])
typeof(df1)
#> [1] "list"

attributes(df1)
#> $names
#> [1] "x" "y"
#> 
#> $class
#> [1] "data.frame"
#> 
#> $row.names
#> [1] 1 2 3
```
A data frame is different from a list in that the length of each of its vectors must be the same. They share properties of both matrices and lists:

A data frame has rownames()30 and colnames(). The names() of a data frame are the column names.

A data frame has nrow() rows and ncol() columns. The length() of a data frame gives the number of columns.

tibble: modern reimagining of a data frame, provided by tibble package
Tibbles are different from data frames in that the class vector is longer and includes tbl_df. 

```{r}
library(tibble)

df2 <- tibble(x = 1:3, y = letters[1:3])
typeof(df2)
#> [1] "list"

attributes(df2)
#> $names
#> [1] "x" "y"
#> 
#> $row.names
#> [1] 1 2 3
#> 
#> $class
#> [1] "tbl_df"     "tbl"        "data.frame"
```
3.6.1 Creating

data.frame(): use to create data frame by supplying name-vector pairs

```{r}
df <- data.frame(
  x = 1:3, 
  y = c("a", "b", "c")
)
str(df)
#> 'data.frame':    3 obs. of  2 variables:
#>  $ x: int  1 2 3
#>  $ y: chr  "a" "b" "c"
```
Beware of the default conversion of strings to factors. Use stringsAsFactors = FALSE to suppress this and keep character vectors as character vectors:

```{r}
df1 <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE
)
str(df1)
#> 'data.frame':    3 obs. of  2 variables:
#>  $ x: int  1 2 3
#>  $ y: chr  "a" "b" "c"
```
Create a tibble similarly to a data frame. The main difference is that tibbles never coerce their input.

```{r}
df2 <- tibble(
  x = 1:3, 
  y = c("a", "b", "c")
)
str(df2)
#> tibble [3 × 2] (S3: tbl_df/tbl/data.frame)
#>  $ x: int [1:3] 1 2 3
#>  $ y: chr [1:3] "a" "b" "c"
```
Data frames automatically transform non-syntactic names (unless check.names = FALSE), tibbles do not.

```{r}
names(data.frame(`1` = 1))
#> [1] "X1"

names(tibble(`1` = 1))
#> [1] "1"
```
While every element of a data frame (or tibble) must have the same length, both data.frame() and tibble() will recycle shorter inputs. However, while data frames automatically recycle columns that are an integer multiple of the longest column, tibbles will only recycle vectors of length one.

```{r}
data.frame(x = 1:4, y = 1:2)
#>   x y
#> 1 1 1
#> 2 2 2
#> 3 3 1
#> 4 4 2
data.frame(x = 1:4, y = 1:3)
#> Error in data.frame(x = 1:4, y = 1:3): arguments imply differing number of
#> rows: 4, 3

tibble(x = 1:4, y = 1)
#> # A tibble: 4 x 2
#>       x     y
#>   <int> <dbl>
#> 1     1     1
#> 2     2     1
#> 3     3     1
#> 4     4     1
tibble(x = 1:4, y = 1:2)
#> Error: Tibble columns must have compatible sizes.
#> * Size 4: Existing data.
#> * Size 2: Column `y`.
#> ℹ Only values of size one are recycled.
```
As opposed to data frames, tibble() allows you to refer to variables created during construction.

```{r}
tibble(
  x = 1:3,
  y = x * 2
)
#> # A tibble: 3 x 2
#>       x     y
#>   <int> <dbl>
#> 1     1     2
#> 2     2     4
#> 3     3     6
```

3.6.2 Row Names

Data frames allow you to label each row with a name, a character vector containing only unique values:

```{r}
df3 <- data.frame(
  age = c(35, 27, 18),
  hair = c("blond", "brown", "black"),
  row.names = c("Bob", "Susan", "Sam")
)
df3
#>       age  hair
#> Bob    35 blond
#> Susan  27 brown
#> Sam    18 black
```
You can get and set row names with rownames() and you can se them to sebset rows.

```{r}
rownames(df3)
#> [1] "Bob"   "Susan" "Sam"

df3["Bob", ]
#>     age  hair
#> Bob  35 blond
```
Row names arise naturally if you think of data frames as 2D structures like matrices: columns (variables) have names so rows (observations) should too. Most matrices are numeric, so having a place to store character labels is important. But this analogy to matrices is misleading because matrices possess an important property that data frames do not: they are transposable. In matrices the rows and columns are interchangeable, and transposing a matrix gives you another matrix (transposing again gives you the original matrix). With data frames, however, the rows and columns are not interchangeable: the transpose of a data frame is not a data frame.

There are three reasons why row names are undesirable:

Metadata is data, so storing it in a different way to the rest of the data is fundamentally a bad idea. It also means that you need to learn a new set of tools to work with row names; you can’t use what you already know about manipulating columns.

Row names are a poor abstraction for labelling rows because they only work when a row can be identified by a single string. This fails in many cases, for example when you want to identify a row by a non-character vector (e.g. a time point), or with multiple vectors (e.g. position, encoded by latitude and longitude).

Row names must be unique, so any duplication of rows (e.g. from bootstrapping) will create new row names. If you want to match rows from before and after the transformation, you’ll need to perform complicated string surgery.

```{r}
df3[c(1, 1, 1), ]
#>       age  hair
#> Bob    35 blond
#> Bob.1  35 blond
#> Bob.2  35 blond
```
For these reasons, tibbles do not support row names. Instead the tibble package provides tools to easily convert row names into a regular column with either rownames_to_column(), or the rownames argument in as_tibble():

```{r}
as_tibble(df3, rownames = "name")
#> # A tibble: 3 x 3
#>   name    age hair 
#>   <chr> <dbl> <chr>
#> 1 Bob      35 blond
#> 2 Susan    27 brown
#> 3 Sam      18 black
```
3.6.3 Printing

One of the most obvious differences between tibbles and data frames is how they print. I assume that you’re already familiar with how data frames are printed, so here I’ll highlight some of the biggest differences using an example dataset included in the dplyr package:

```{r}
dplyr::starwars
#> # A tibble: 87 x 14
#>    name  height  mass hair_color skin_color eye_color birth_year sex   gender
#>    <chr>  <int> <dbl> <chr>      <chr>      <chr>          <dbl> <chr> <chr> 
#>  1 Luke…    172    77 blond      fair       blue            19   male  mascu…
#>  2 C-3PO    167    75 <NA>       gold       yellow         112   none  mascu…
#>  3 R2-D2     96    32 <NA>       white, bl… red             33   none  mascu…
#>  4 Dart…    202   136 none       white      yellow          41.9 male  mascu…
#>  5 Leia…    150    49 brown      light      brown           19   fema… femin…
#>  6 Owen…    178   120 brown, gr… light      blue            52   male  mascu…
#>  7 Beru…    165    75 brown      light      blue            47   fema… femin…
#>  8 R5-D4     97    32 <NA>       white, red red             NA   none  mascu…
#>  9 Bigg…    183    84 black      light      brown           24   male  mascu…
#> 10 Obi-…    182    77 auburn, w… fair       blue-gray       57   male  mascu…
#> # … with 77 more rows, and 5 more variables: homeworld <chr>, species <chr>,
#> #   films <list>, vehicles <list>, starships <list>
```

Tibbles only show the first 10 rows and all the columns that will fit on screen. Additional columns are shown at the bottom.

Each column is labelled with its type, abbreviated to three or four letters.

Wide columns are truncated to avoid having a single long string occupy an entire row. (This is still a work in progress: it’s a tricky tradeoff between showing as many columns as possible and showing columns in their entirety.)

When used in console environments that support it, colour is used judiciously to highlight important information, and de-emphasise supplemental details.

3.6.4 Subsetting

In my opinion, data frames have two undesirable subsetting behaviours:

When you subset columns with df[, vars], you will get a vector if vars selects one variable, otherwise you’ll get a data frame. This is a frequent source of bugs when using [ in a function, unless you always remember to use df[, vars, drop = FALSE].

When you attempt to extract a single column with df$x and there is no column x, a data frame will instead select any variable that starts with x. If no variable starts with x, df$x will return NULL. This makes it easy to select the wrong variable or to select a variable that doesn’t exist.

-Tibbles tweak these behaviours so that a [ always returns a tibble, and a $ doesn’t do partial matching and warns if it can’t find a variable (this is what makes tibbles surly).

```{r}
df1 <- data.frame(xyz = "a")
df2 <- tibble(xyz = "a")

str(df1$x)
#>  chr "a"
str(df2$x)
#> Warning: Unknown or uninitialised column: `x`.
#>  NULL
```
A tibble’s insistence on returning a data frame from [ can cause problems with legacy code, which often uses df[, "col"] to extract a single column. If you want a single column, I recommend using df[["col"]]. This clearly communicates your intent, and works with both data frames and tibbles.

3.6.5 Testing and Coercing

is.data.frame(): use to test if an object is a data frame or a tibble

```{r}
is.data.frame(df1)
#> [1] TRUE
is.data.frame(df2)
#> [1] TRUE
```

Typically, it should not matter if you have a tibble or data frame, but if you need to be certain, use is_tibble():

```{r}
is_tibble(df1)
#> [1] FALSE
is_tibble(df2)
#> [1] TRUE
```

as.data.frame(): coerce to a data frame
as_tibble(): coerce to a tibble

3.6.6 List Columns

Since a data frame is a list of vectors, it is possible for a data frame to have a column that is a list. This is very useful because a list can contain any other object: this means you can put any object in a data frame. This allows you to keep related objects together in a row, no matter how complex the individual objects are.

List-columns are allowed in data frames but you have to do a little extra work by either adding the list-column after creation or wrapping the list in I()32.

```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)

data.frame(
  x = 1:3, 
  y = I(list(1:2, 1:3, 1:4))
)
#>   x          y
#> 1 1       1, 2
#> 2 2    1, 2, 3
#> 3 3 1, 2, 3, 4
```

List columns are easier to use with tibbles because they can be directly included inside tibble() and they will be printed tidily:

```{r}
tibble(
  x = 1:3, 
  y = list(1:2, 1:3, 1:4)
)
#> # A tibble: 3 x 2
#>       x y        
#>   <int> <list>   
#> 1     1 <int [2]>
#> 2     2 <int [3]>
#> 3     3 <int [4]>
```

3.6.7 Matrix and Data Frame Columns

As long as the number of rows matches the data frame, it’s also possible to have a matrix or array as a column of a data frame. (This requires a slight extension to our definition of a data frame: it’s not the length() of each column that must be equal, but the NROW().) As for list-columns, you must either add it after creation, or wrap it in I().

```{r}
dfm <- data.frame(
  x = 1:3 * 10
)
dfm$y <- matrix(1:9, nrow = 3)
dfm$z <- data.frame(a = 3:1, b = letters[1:3], stringsAsFactors = FALSE)

str(dfm)
#> 'data.frame':    3 obs. of  3 variables:
#>  $ x: num  10 20 30
#>  $ y: int [1:3, 1:3] 1 2 3 4 5 6 7 8 9
#>  $ z:'data.frame':   3 obs. of  2 variables:
#>   ..$ a: int  3 2 1
#>   ..$ b: chr  "a" "b" "c"
```

Matrix and data frame columns require a little caution. Many functions that work with data frames assume that all columns are vectors. Also, the printed display can be confusing.

```{r}
dfm[1, ]
#>    x y.1 y.2 y.3 z.a z.b
#> 1 10   1   4   7   3   a
```

3.6.8 Exercises

1. Can you have a data frame with zero rows? What about zero columns?

2. What happens if you attempt to set rownames that are not unique?

3. If df is a data frame, what can you say about t(df) and t(t(df))? Perform some experiments, making sure to try different column types.

4. What does as.matrix() do when applied to a data frame with columns of different types? How does it differ from data.matrix()?

3.7 NULL

NULL is special because it has a unique type, its length is always 0, and it can't have any attributes.

```{r}
typeof(NULL)
#> [1] "NULL"

length(NULL)
#> [1] 0

x <- NULL
attr(x, "y") <- 1
#> Error in attr(x, "y") <- 1: attempt to set an attribute on NULL
```

is.null(): use to test for NULL

```{r}
is.null(NULL)
#> [1] TRUE
```

There are two common uses of NULL:

To represent an empty vector (a vector of length zero) of arbitrary type. For example, if you use c() but don’t include any arguments, you get NULL, and concatenating NULL to a vector will leave it unchanged:

```{r}
c()
#> NULL
```

To represent an absent vector. For example, NULL is often used as a default function argument, when the argument is optional but the default value requires some computation (see Section 6.5.3 for more on this). Contrast this with NA which is used to indicate that an element of a vector is absent.

3.8 QUIZ ANSWERS

1. The four common types of atomic vector are logical, integer, double and character. The two rarer types are complex and raw.

2. Attributes allow you to associate arbitrary additional metadata to any object. You can get and set individual attributes with attr(x, "y") and attr(x, "y") <- value; or you can get and set all attributes at once with attributes().

3. The elements of a list can be any type (even a list); the elements of an atomic vector are all of the same type. Similarly, every element of a matrix must be the same type; in a data frame, different columns can have different types.

4. You can make a list-array by assigning dimensions to a list. You can make a matrix a column of a data frame with df$x <- matrix(), or by using I() when creating a new data frame data.frame(x = I(matrix())).

5. Tibbles have an enhanced print method, never coerce strings to factors, and provide stricter subsetting methods.
